---
title: "Loan Repayment and borrower analysis"
output: 
  html_document:
        code_folding: hide
        toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE, message=FALSE}
# Loading Libraries
libs <- c("dplyr", "tidyr", "data.table", "ggplot2", "caret", "caTools", "car", "MASS", "gridExtra", "broom", "DT")
lapply(libs, require, character.only = T)
```
## The Loan Data
```{r loading data, message=FALSE}
# loading data
data <- fread("data/loan_data.csv")
data <-  data %>%
        mutate(not.fully.paid = factor(ifelse(not.fully.paid == 1, "Yes", "No")),
               credit.policy = factor(ifelse(credit.policy == 1, "Yes", "No")),
               annual.inc = exp(log.annual.inc),
               days = round(data$days.with.cr.line))

```

| Variable  | class     | description                    |
|:----------|:----------|:-------------------------------|
| credit_policy   | numeric | 1 if the customer meets the credit underwriting criteria; 0 otherwise. |
| purpose      | character | The purpose of the loan. |
| int_rate      | numeric   | The interest rate of the loan (more risky borrowers are assigned higher interest rates).  |
| installment  | numeric   | The monthly installments owed by the borrower if the loan is funded. |
| log_annual_inc | numeric   | The natural log of the self-reported annual income of the borrower. |
| dti | numeric   | The debt-to-income ratio of the borrower (amount of debt divided by annual income). |
| fico | numeric   | The FICO credit score of the borrower. |
| days_with_cr_line | numeric   | The number of days the borrower has had a credit line. |
| revol_bal | numeric   | The borrower's revolving balance (amount unpaid at the end of the credit card billing cycle). |
| revol_util | numeric   | The borrower's revolving line utilization rate (the amount of the credit line used relative to total credit available). |
| inq_last_6mths | numeric   | The borrower's number of inquiries by creditors in the last 6 months. |
| delinq_2yrs | numeric   | The number of times the borrower had been 30+ days past due on a payment in the past 2 years. |
| pub_rec | numeric   | The borrower's number of derogatory public records.
| not_fully_paid | numeric   | 1 if the loan is not fully paid; 0 otherwise.

## Exploratory Data Analysis
### Exploring Possible response variables
#### Examining fully paid 
```{r, message=FALSE}
data %>%
ggplot(aes(x = factor(not.fully.paid), fill = not.fully.paid)) +
  geom_bar() +
  labs(x = "Default", y = "Count", title = "Number of Clients who did and did not Default") +
  theme_minimal() +
  theme(legend.position = "none")

fully_paid_prop <- data %>%
  summarize(prop = mean(not.fully.paid == "No")) %>%
  pull()

Counts <- data %>%
  count(not.fully.paid)
```

  When looking at the number of people who defaulted (did not completely pay off the loan) on their loan, there is a significant difference between the number of individuals who have paid off their loans completely vs those who defaulted on it. With around `r fully_paid_prop*100`% (`r Counts[1,2]`) paying off their loans completely and around `r (1-fully_paid_prop)*100`% (`r Counts[2,2]`) defaulting on their loans, this as a response variable is unbalanced and will require balancing during the prediction process. 


#### Examining days with credit line
```{r, message=FALSE}
days_his1 <- data %>% 
  ggplot(aes(x = days.with.cr.line, color = factor(not.fully.paid))) +
  geom_histogram() +
  labs(x = "Days with Credit Line", y = "Count", title = "Histogram of Days with Credit Line by Defaults", color = "Default") +
  theme_minimal()

days_his2 <- data %>% 
  ggplot(aes(x = days.with.cr.line, color = factor(purpose))) +
  geom_histogram() +
  labs(x = "Days with Credit Line", y = "Count", title = "Histogram of Days with Credit Line by Purpose", color = "Purpose") +
  scale_color_discrete(labels = c("Other", "Credit Card", "Debt Consolidation", "Educational", "Home Improvement", "Major Purchase", "Small Business")) +
  theme_minimal()

grid.arrange(days_his1, days_his2)
```

Based on the histograms above, the distribution of days with a credit line stays relatively identical across the different populations when accounting for either defaults or by the purpose of the loans. Moreover, as the population increases, within the groups, it appears to approach a right-tail skewed normal distribution.

### Delinquency
```{r, message=FALSE}
# Public Record
ggplot(data, aes(factor(pub.rec), fico, fill = factor(pub.rec))) +
  geom_boxplot() +
  labs(x = "Number of derogatory Public Records", y = "FICO Score", title = "Box Plot of number of FICO against the number of derogatory Public Records") +
  theme_minimal() +
  theme(legend.position = "none")

# Delinquencies
ggplot(data, aes(factor(delinq.2yrs), int.rate, fill = factor(delinq.2yrs))) +
  geom_boxplot() +
  labs(x = "Number of Delinquencies in 2 years", y = "Interest Rate", title = "Box Plot of number of interest rate against the Number of Delinquencies in 2 years") +
  theme_minimal() +
  theme(legend.position = "none")
  
ggplot(data, aes(factor(delinq.2yrs), log.annual.inc, fill = factor(delinq.2yrs))) +
  geom_boxplot() +
  labs(x = "Number of Delinquencies in 2 years", y = "Interest Rate", title = "Box Plot of number of FICO against the Number of Delinquencies in 2 years") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Inferential Modeling
```{r}
data %>% 
  ggplot(aes(x = fico, y = days.with.cr.line, color = factor(not.fully.paid))) +
  geom_point() +
  geom_smooth(method = "lm")

data %>%
  ggplot(aes(x = fico, y = int.rate, color = factor(not.fully.paid))) +
  geom_point() +
  geom_smooth(method = "lm")

data %>% 
  ggplot(aes(x = fico, y = revol.util, color = not.fully.paid)) +
  geom_point() +
  geom_smooth(method = "lm")

Simple_ficos <- lm(cbind(days.with.cr.line, int.rate, revol.util) ~ fico + not.fully.paid, data = data)
datatable(tidy(Simple_ficos))
```


```{r}
data %>%
  ggplot(aes(x = log.annual.inc, y = fico)) +
  geom_point() +
  geom_smooth(method = "lm")
```

##
```{r}
data %>%
  ggplot(aes(x = log.annual.inc, y = days.with.cr.line, color = not.fully.paid)) +
  geom_point() +
  geom_smooth(method = "lm")
```


# Model Analysis
## Determining factors contributing to a person defaulting on the loan
```{r}
summary(complete_binomial <- glm(formula = not.fully.paid ~ . - days - annual.inc, data = data, family = "binomial"))
Stepwise_binomial <- stepAIC(complete_binomial, direction = "both")
summary(Stepwise_binomial)
vif(Stepwise_binomial)
```



## Determining factors contributing to days with the credit line
```{r}
lm(days.with.cr.line ~ fico, data = data) %>% summary()
lm(days.with.cr.line ~ log.annual.inc + poly(fico, 3), data = data) %>% summary()
complete_days <- lm(days.with.cr.line ~ .- days - annual.inc - not.fully.paid, data = data)
forward_selection <- stepAIC(complete_days, direction = "forward")
summary(forward_selection)
vif(forward_selection)
```

## Gamma Regression
### Estimating Days with credit line based on character traits
```{r}
# Variables of interest: 
##
plot(log(days) ~ fico, data = data)
plot((1/days) ~ fico, data = data)
summary(pois_reg <- glm(days ~ int.rate*fico + I(int.rate^2), data = data, family = Gamma(link = "log")))
summary(pois_reg <- glm(days ~ . + I(int.rate^2), data = data, family = Gamma(link = "log")))
test_data <- data[runif(500, min = 0, max = nrow(data)),]
pred <- predict(pois_reg, newdata = test_data, type = "response")
actual <- test_data[,days.with.cr.line]

pchisq(pois_reg$deviance, pois_reg$df.residual, lower.tail=F)

plot(pois_reg)
plot(pred, actual)
plot(log(pred), actual)
hist(pred)
hist(actual)

summary(lm(actual ~ log(pred)))
```

### Gamma Regression Model Stacking (Prediction Oriented)
```{r}



```



# Cluster Analysis and Dimensionality reduction
```{r}
# summary(pr <- prcomp(data[,c("fico", "revol.util", "days.with.cr.line", "dti")], scale = T))
# biplot(pr)
# km1 <- kmeans(scaled, centers = 3, nstart = 1000)
# plot(data$fico, data$days.with.cr.line , col = km1$cluster)
```

```{r}
# (pr <- prcomp(data[,c(3,6,7,9,10)], scale = T))
# summary(pr)
# biplot(pr)
# plot(pr$x[,c(1,2)], col = as.factor(data$not.fully.paid))
# plot(pr$x[,c(1,3)], col = as.factor(data$not.fully.paid))
# plot(pr$x[,c(2,3)], col = as.factor(data$not.fully.paid))
# 
# km1 <- kmeans(pr$x[,1:3], centers = 3, nstart = 1000)
# plot(pr$x[,1:2], col = km1$cluster)
# table(km1$cluster, data$not.fully.paid)
```

```{r}
# (pr <- prcomp(data[,c(3,6,7,9,10)], scale = T))
# summary(pr)
# biplot(pr)
# plot(pr$x[,c(1,2)], col = as.factor(data$not.fully.paid))
# plot(pr$x[,c(1,3)], col = as.factor(data$not.fully.paid))
# plot(pr$x[,c(2,3)], col = as.factor(data$not.fully.paid))
# 
# km1 <- kmeans(pr$x[,1:3], centers = 3, nstart = 1000)
# plot(pr$x[,1:2], col = km1$cluster)
# table(km1$cluster, data$not.fully.paid)
```


# Machine Learning Prediction Models using Caret
## 
```{r}
# set.seed(0)
# ind <- sample(nrow(data), size = round(nrow(data)*.8))
# train <- data[ind,]
# test <- data[-ind,]
# 
# 
# formula = 
# 
# ctrl <- trainControl(method = "repeatedcv", repeats = 5,
#                      classProbs = TRUE,
#                      summaryFunction = twoClassSummary,
#                      sampling = "down")
# 
# gbm <- train(formula,
#              data = train,
#              method = "gbm",
#              metric = "ROC",
#              verbose = F,
#              trControl = ctrl) 
# 
# probs <- predict(gbm, test, type = "prob")
# preds <- ifelse(probs[1] > .50, "No", "Yes")
# table(preds, test$credit.policy)
```


## 